Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions

### 2021-03-05 Latex/appendix/webappendix.Rnw
txt update

\begin{table}[h!]
\footnotesize
\centering
\caption{GEO sample accession codes of the analyzed data from the ENCODE Consortium}
\begin{tabular}{lccccccc}
\hline
Cell Line & H3K27me3 & H3K36me3 & EZH2 & H3K4me3 & H3K27ac & CTCF & RNA-seq\\
\hline
H1hesc & GSM733748 & GSM733725 & GSM1003524 & GSM733657 & GSM733718 & GSM733672 & GSM758566\\
HelaS3 & GSM733696 & GSM733711 & GSM1003520 & GSM733682 & GSM733684 & GSM733785 & GSM765402\\
Hepg2 & GSM733754 & GSM733685 & GSM1003487 & GSM733737 & GSM733743 & GSM733645 & GSM758575\\
Huvec & GSM733688 & GSM733757 & GSM1003518 & GSM733673 & GSM733691 & GSM733716 & GSM758563\\
\hline
\end{tabular}
\end{table}

@@

>>>>

\begin{table}[h!]
\tiny
\centering
\caption{GEO sample accession codes of the analyzed data from the ENCODE Consortium}
\begin{tabular}{lcccccccc}
\hline
Cell Line & Control & H3K27me3 & H3K36me3 & EZH2 & H3K4me3 & H3K27ac & CTCF & RNA-seq\\
\hline
H1hesc & GSM733770 & GSM733748 & GSM733725 & GSM1003524 & GSM733657 & GSM733718 & GSM733672 & GSM758566\\
HelaS3 & GSM733659 & GSM733696 & GSM733711 & GSM1003520 & GSM733682 & GSM733684 & GSM733785 & GSM765402\\
Hepg2 & GSM733732 & GSM733754 & GSM733685 & GSM1003487 & GSM733737 & GSM733743 & GSM733645 & GSM758575\\
Huvec & GSM733715 & GSM733688 & GSM733757 & GSM1003518 & GSM733673 & GSM733691 & GSM733716 & GSM758563\\
\hline
\end{tabular}
\end{table}

@@

All ENCODE data analysis presented in the main text, and in this Web Appendix, were adjusted for input control samples in methods that are designed to do so. Specifically, differential peaks from ChIPComp, DiffBind, diffReps, and THOR were obtained on ChIP-seq samples while passing input control samples in their software call as an optional input argument argument. In contrast, input control samples were not considered in epigraHMM, csaw, and RSEG analyses.

### 2021-03-05 Latex/letter/letter.Rmd
txt update

---
title: "Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'"
author: "Pedro L. Baldoni, Naim U. Rashid, Joseph G. Ibrahim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

@@

@@

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested, we add the following sentence to our real data analysis section 'We compared the genome-wide performance of the presented model with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithmic approaches for both broad and short ChIP-seq enrichment profiles.'. We believe that such a list covers the most used algorithms for DPC in the field of bioinformatics.

@@

> We acknowledge the issue of GC-content bias in the referenced paragraph. We cite the work of Teng and Irizarry (2017) that proposed a correction method for GC-content bias in ChIP-seq analysis. Their method is applicable to any peak calling algorithm that measures the enrichment of reads in bins, such as our method. Later in the article, we point out to the reader that our method is able to account for GC-content bias in the same manner as suggested by Teng and Irizarry (2017) by using their count-normalizing denominators in our model as offsets. We describe this process in detail in Section B1.3 of the Web Appendix.

@@

> This has been addressed. In the second paragraph of Section 5, we reference the current GitHub repository of MACS2 \url{https://github.com/macs3-project/MACS}.

@@

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution and did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), a method that was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyze epigenomic marks with broad enrichment. A possible explanation for such a lack of effect of GC-content on the analyzed broad ChIP-seq data is that, in such scenarios, the enrichment of counts has a much lower signal-to-noise ratio with respect to background regions in comparison to short epigenomic marks. In addition, the high resolution of conts computed on non-overlapping windows of 500bp makes the effect of GC-content on enrichment regions indistinguishable from the GC-content effect on background regions. We shown in Web Figure 7 results from our GC-content analysis utizilin the epigenomic mark EZH2. Due to these reasons, we did not explore alternative methods for GC-bias correction for broad marks.

@@

> During the $t^{th}$ iteration of the EM-algorithm, one needs to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximize $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$ during the maximization step, such that

@@

$$ Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mathbf{y}_{..j},\mathbf{x};\boldsymbol{\Psi}^{(t)})\delta_{l}\log f_{(2,l)}\left(\mathbf{y}_{..j}|\mathbf{x}_l;\boldsymbol{\psi}_{(2,l)}\right),\text{ and,}$$
$$ Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{3}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{3}\right),$$

> where the notation is precisely defined in Section 3 of the main text. Note that $Q_{1}$, $Q_{2}$, and $Q_{3}$ above can all be seen as weighted NB regression models where the weights are formed by a combination of HMM posterior probabilties and/or mixture model posterior proportions/probabilities. Given the chosen constrained parametrization across all HMM states, we use numerical gradient-based optimization to obtain the parameter estimates. In the equations above, $\mathbf{y}_{..j}$ is the sequence of read counts observed across all conditions and replicates on genomic window $j$. On the other hand, the design matrix $x$ is a a matrix of zeros and ones of dimention $G\times L$ that encodes all differential combinatorial patterns of enrichment $L$ across $G$ conditions (see Figure 1 of the main paper for an example). Further details of the implemented RCEM algorithm are detailed in Web Appendix B4.

@@

> We thank the reviewer for pointing this out. We believe that comparing the relative bias of parameter estimates across different SNR levels is difficult because different SNRs are associated with different parameter values (some larger than others) in our simulation study. Although we did observe an improvement in performance in terms of model sensitivity/specificity of peak calls in higher SNR data sets, the main driving quantity for the reduction of parameter bias is the number of replicates included in our model. For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the observed relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was particularly significant in scenarios with four conditions with respect to parameters $\beta_{3}$ and $\lambda_{3}$. In scenarios with higher number of conditions, these results highlight the importance of experimental replicates to achieve precise parameter estimates.". The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K27me3. This is now reported in our manuscript. We present the estimate SNR value for other scenarios in our Web Appendix (Web Tables 2-3)

@@

@@

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper. For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed. Also, note that the strategy of utilizing ChIP-seq data to define sets of silver-standard marks for benchmarking purposes has been already used by others (Ji et. al, 2013; PNAS). Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer.

> NAIM, PLEASE SEE MY REPORT IN THE REPO UNDER ./Supp/Report_RNA_vs_ChIP/ FOR A COMPREHENSIVE DISCUSSION ABOUT HOW USING RNA-SEQ TO DEFINE DE GENES AND THEN BENCHMARKING DIFFERENTIAL H3K36ME3 CHIP-SEQ PEAKS CAN BE PROBLEMATIC. WE DO NOT BENCHMARK DIFFERENTIAL PEAK CALLING METHODS IN THE PAPER USING RNA-SEQ. AS I MENTION IN THE PAPER:  'Following the analysis of \cite{ji2013differential}, we defined a set of protein coding genes exhibiting $|\text{LFC}|>2$ of ChIP-seq counts between Helas3 and Hepg2 cell lines as true differential binding sites. Results using different threshold levels are presented in the Web Appendix D1 and agree with those presented here.'. I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER). THE ONLY PART OF THE PAPER WHERE WE USE RNA-SEQ IS IN CHROMATIN SEGMENTATION (FIGURE 5).

@@

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 were associated with differential enrichment for the H3K36me3 mixture model component. That is, genes that are expected to be actively transcribed in Helas3 (because they are marker genes) were associated with differential peaks and, in addition, were associated with the differential combinatorial pattern of enrichment for H3K36me3 alone. Note that consensus enrichment is a rare state and not expected to be observed in this analysis. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. That is, for each marker gene and the overlapping differential peaks, we calculate the average window-based mixture proportion from the differential combinatorial pattern associated with enrichment for H3K36me3 alone. Note that our selected model had two mixture components in the differential HMM state (either enrichment of H3K36me3 alone, or enrichment of H3K27me3 and EZH2 together). These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

@@

> I have changed "To quantify local genome activity" to "To quantify local epigenomic activity on a genome-wide scale".

@@

@@

@@

> This has been addressed and made clear in text. For RSEG, the metrics were calculated based on the instances where it ran without issues (71 out of 100 simulations).

@@

> This has been addressed. The tables and figures of the supp. are already indexed with 'Web' in front of it (e.g. Web Figure 1).

@@

> All analyses in the paper were conducted using a single CPU per task without multithreading. Figures 4E and 3E now exhibit the peak memory comsumption by each method along with the total computing time. We observed that in broad marks (500bp) our method was among the most efficient ones, requiring only 5GB of memory and 30 min for a genome-wide analysis of 4 samples (2 conditions, 2 replicates per condition). In narrow marks (250bp), our method required nearly 10GB of memory for a computation time of about 1 hour. We believe that these are reasonable quantities and quite comparable to other  efficient algorithms. For example, the widely used method DiffBind required nearly as much memory even though it is a method that relies on peaks called a priori (MACS2 in our analysis) and had a much lower performance in broad histone modification marks.

@@

> As mentioned in our paper, we added a discussion in Web Appendix regarding the model selection via BIC. Web Figures 8-10 include results from our simulation study exploring the presented strategy to choose the optimal number of combinatorial patterns via BIC. Due to the lack of space, we have added the following sentence to our Discussion "In situations where certain combinatorial patterns are rare, prunning of specific combinatorial patterns is possible via model selection via BIC (Web Appendix B2)".

@@

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented in this section pertain to the analysis of two cell lines, namely Helas3 and Hepg2. Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR)". Similarly, input controls were also generated in our simulation study and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance in epigraHMM by accounting for input control effect in differential peak calling (See Web Figure 2 for these results). This effect was also reported by others (Lun \& Smyth, 2015). We did not compare in our paper the performance of other methods with and without accounting for input controls. As stated above, analyses reported in our paper were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR). For all other methods (epigraHMM, RSEG, and csaw), input controls were not considered and their results are reported as such.

@@

> In Figure 2A, the observed FDR is shown for all methods and for all assessed nominal FDR levels, for all methods. The observed FDR shown in Figure 2A is quite close to the nominal FDR chosen (note vertical lines on the panel and corresponding nominal FDR points). In Figure 3A, no FDR was or is currently reported. Instead, we show 1-specificity on the x-axis. In Figure 3B, we show the FDR for all methods. As we can see, the observed and nominal FDR levels reported by our method were quite close to each other (0.05, 0.07, 0.08, 0.09, 0.10). In comparison, RSEG exhibited observed FDR levels of 0.23, 0.63, 0.69, 0.74,0.80 for nominal FDR levels of 0.01, 0.05, 0.10, 0.15, and 0.20. We thank the reviewer for this comment but we do believe our results are superior than the suggested FDR control of RSEG.

@@

> We thank the reviewer for the constructive comment. As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene. In contrast, Figure 3 below presents an additional example of a DE gene and without H3K36me3 enrichment in both Helas3 and Hepg2 cell lines. Moreover, not all the analyzed marks in this paper are associated with the expression level of the nearest gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. In several instances, enrichment of the repressive mark H3K27me3 occurs on inter-genic regions with nearby expressed genes. Figure 4 below shows an example of such a region. Due to these reasons, we believe that the proposed analysis would be suboptimal to the one presented in this paper for the analyzed cell lines.

> NAIM, PLEASE SEE MY COMMENT ABOVE ABOUT THIS GENE EXPRESSION ANALYSIS. I CITE THE PNAS PAPER FROM HONGKAI JI (2013), WHERE THEY DEFINED GOLD-STANDARD BENCHMARKING MARKS BASED ON LOG FOLD CHANGES OF CHIP-SEQ COUNTS, NOT ON RNA-SEQ COUNTS (SEE PAGE 6971 OF THEIR PAPER). PLEASE, SEE MY REPORT UNDER ./Supp/Report_RNA_vs_ChIP/ AND LET ME KNOW IF ANY OF THOSE RESULTS SHOULD BE INCLUDED HERE TO SUPPORT OUR RESPONSES.

@@

> As we discuss above, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential enrichment for the H3K36me3 mixture model component. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior proportion associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

>>>>

---
title: "Response Letter for Manuscript ID BIOM2020344P - 'Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions'"
author: "Pedro L. Baldoni, Naim U. Rashid, Joseph G. Ibrahim"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
urlcolor: blue
---

@@

Dear reviewers,

We thank you for the constructive comments regarding our paper "Efficient Detection and Classification of Epigenomic Changes Under Multiple Conditions".

Our revised paper has 20 pages in total, as requested, including appendices and references, and excluding the title page and abstract, figures and tables. We emphasize that the our presented method, previously called mixNBHMM, is now called epigraHMM. We refer to our method as epigraHMM throughout this revised version of our paper.

Detailed point-by-point responses to all comments are provided below.

@@

> We thank the reviewer for this constructive comment. We cite in our paper the benchmarking article 'A comprehensive comparison of tools for differential ChIP-seq analysis' (Briefings in Bioinformatics, 2016). However, we are not aware of a more recent comprehensive review of DPCs algorithms in the literature. As suggested, we add the following sentence to our real data analysis section 'Here, we compared epigraHMM with the widely used DPCs ChIPComp, csaw, DiffBind, diffReps, RSEG, and THOR, a list that covers a variety of algorithms for both broad and short ChIP-seq enrichment profiles (...)'. We believe that such a list covers the most used algorithms for DPC in the field of bioinformatics.

@@

> We acknowledge the issue of GC-content bias in the referenced paragraph. We cite the work of Teng and Irizarry (2017) that proposed a correction method for GC-content bias in ChIP-seq analysis. Their method is applicable to any peak calling algorithm that measures the enrichment of reads in bins, such as our method epigraHMM. Later in the article, we point out to the reader that our method is able to account for GC-content bias in the same manner as suggested by Teng and Irizarry (2017) by using their count-normalizing denominators in our model as offsets. We describe this process in detail in Section B1.3 of the Web Appendix.

@@

> This has been addressed. As requested by the Editorial Office, we reference the current GitHub repository of MACS2 \url{https://github.com/macs3-project/MACS} in the Supporting Information section of our revised paper.

@@

> As mentioned above, our method is able to account go GC-bias via model offsets. These model offsets can be obtained by methods such as the one proposed by Teng and Irizarry (2017). We present a simulation study in Section B1.3 of the Web Appendix showing that epigraHMM leads to higher sensitivity/specificity when accounting for GC-bias even after read count normalization using the count-normalizing denominators from Teng and Irizarry (2017). However, for the broad marks that we analyzed in our manuscript, we did not observe a substantial effect of GC-content on read count distribution. Also, we did not observe any substantial improvement in accounting for GC-content in our model via offsets when utilizing the method from Teng and Irizarry (2017), which was developed only for narrow marks characterized by sharp peaks. In fact, the authors do not recommend the use of their method when analyzing epigenomic marks with broad enrichment. A possible explanation for such a lack of effect of GC-content on the analyzed broad ChIP-seq data is that, in such scenarios, the enrichment of counts has a much lower signal-to-noise ratio with respect to background regions in comparison to short epigenomic marks. In addition, the high resolution of conts computed on non-overlapping windows of 500bp may lead to a non-differential effect of GC-content on enrichment regions in comparison to the GC-content effect on background regions. We shown in Web Figure 7 results from our GC-content analysis utilizing the epigenomic mark EZH2. Due to these reasons, we did not explore alternative methods for GC-bias correction for broad marks.

@@

> During the $t^{th}$ iteration of the EM-algorithm, one needs to obtain $\beta_{1}^{(t)}, \beta_{3}^{(t)}, \lambda_{1}^{(t)}$, and $\lambda_{3}^{(t)}$ that maximizes $Q_{1}\left(\boldsymbol{\psi}_{1}|\boldsymbol{\Psi}^{(t)}\right) + Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) + Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right)$ during the maximization step, where

@@

$$ Q_{2}\left(\boldsymbol{\delta},\boldsymbol{\psi}_{2}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=2|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\sum_{l=1}^{L}Pr(W_{jl}=1|Z_{j}=2,\mathbf{y}_{..j},\mathbf{x};\boldsymbol{\Psi}^{(t)})\log \delta_{l}f_{(2,l)}\left(\mathbf{y}_{..j}|\mathbf{x}_l;\boldsymbol{\psi}_{(2,l)}\right),\text{ and,}$$
$$ Q_{3}\left(\boldsymbol{\psi}_{3}|\boldsymbol{\Psi}^{(t)}\right) = \sum_{j=1}^{M}Pr\left(Z_{j}=3|\mathbf{y},\mathbf{x};\boldsymbol{\Psi}^{(t)}\right)\log f_{3}\left(\mathbf{y}_{..j}|\boldsymbol{\psi}_{3}\right),$$

> where the notation is precisely defined in Section 3 of the main text. Note that $Q_{1}$, $Q_{2}$, and $Q_{3}$ above can all be seen as weighted NB regression models where the weights are formed by HMM marginal posterior probabilties ($Q_{1}$ and $Q_{3}$), and a combination of HMM marginal posterior probabilities, mixture model mixture posterior probabilities, and mixture model proportions ($Q_{2}$). Given the chosen constrained parametrization across all HMM states, we use numerical gradient-based optimization to obtain the parameter estimates. In the equations above, $\mathbf{y}_{..j}$ is the sequence of read counts observed across all conditions and replicates on genomic window $j$. On the other hand, the design matrix $x$ is a a matrix of zeros and ones of dimention $G\times L$ that encodes all differential combinatorial patterns of enrichment $L$ across $G$ conditions (see Figure 1 of the main paper for an example). Further details of the implemented RCEM algorithm are detailed in Web Appendix B4.

@@

> We thank the reviewer for pointing this out. We believe that comparing the relative bias of parameter estimates across different SNR levels is difficult because different SNRs are associated with different parameter values (some larger than others) in our simulation study. Although we did observe an improvement in performance in terms of model sensitivity/specificity of peak calls in higher SNR data sets, the main driving quantity for the reduction of parameter bias is the number of replicates included in our model. For this reason, we rephrase the mentioned sentence in our paper as "Depending on the number of conditions, the relative bias and the range of the reported percentiles tended to decrease as more replicates were included in the analyses. This effect was more pronounced when estimating $\beta_{3}$ and $\lambda_{3}$ in scenarios with four conditions, which highlights the importance of experimental replicates to achieve precise parameter estimates.". The estimated SNR level from ENCODE was 3.21 (ratio of mean counts between peak and background regions) for H3K27me3. This is now reported in our manuscript. We present the estimate SNR value for other scenarios in our Web Appendix (Web Tables 2-3)

@@

> We thank the reviewer for the constructive comment. We prepared a report to support our response to this comment, which can be assessed at https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf

@@

> Regarding the use of 'matched differentially expressed genes as the gold standard to evaluate the differential transcribed genes called from differential H3K36me3 activity', we did not observe a sufficient correspondance between gene expression levels and H3K36me3 enrichment for the cancer cell lines analyzed in our paper (https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf). For instance, we have found several instances where genes were expressed but no enrichment of H3K36me3 was observed. Also, note that the strategy of utilizing ChIP-seq data to define sets of silver-standard marks for benchmarking purposes has been already used by others (see Figure 2E in Ji et. al, 2013; PNAS). Below, we present some of these problematic cases obtained from the UCSC Genome Browser. Note in Figure 1, for example, that both cell lines Helas3 and Hepg2 did not show enrichment for H3K36me3 in the rightmost region of the plot while the associated gene body was clearly expressed. Note in Figure 2, for example, the lack of agreement between the transcription level of the reported gene and the enrichment for H3K36me3 in cell line Hepg2. Due to these reasons, we decided not to use gene expression levels when benchmarking H3K36me3 as suggested by the reviewer. The linked report provides several other problematic cases with the peak call tracks and the entire RNA-seq data analysis.

@@

> Regarding the biological significance of our method, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 were associated with the differential combinatorial pattern associated with enrichment for H3K36me3 alone in epigraHMM's mixture model. That is, genes that are expected to be actively transcribed in Helas3 (because they are marker genes) were associated with differential peaks and, in addition, were associated with the differential combinatorial pattern of enrichment for H3K36me3 alone. Note that consensus enrichment is a rare state and not expected to be observed in this analysis. The presented differential associated enrichment level shown in Figure 5E is the average mixture model posterior probability associated with the mixture component that model the enrichment of H3K36me3 alone. That is, for each marker gene and the overlapping differential peaks, we calculate the average window-based posterior probability from the differential combinatorial pattern associated with enrichment for H3K36me3 alone. Note that our selected model had two mixture components in the differential HMM state (either enrichment of H3K36me3 alone, or enrichment of H3K27me3 and EZH2 together). These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3.

@@

> We have changed "To quantify local genome activity" to "To quantify local epigenomic activity on a genome-wide scale".

@@

> This has been addressed in our paper and we utilize the proper terminology in Section 2.

@@

> This has been addressed in our revised paper.

@@

> This has been addressed and made clear in text. For RSEG, the metrics were calculated based on the instances where it ran without issues. In other words, we did not consider problematic cases from RSEG when computing metrics of sensitivity, FDR, computing time, etc. in our simulation study.

@@

> This has been addressed. The tables and figures of the Supplementary Information file are indexed with 'Web' in front of it (e.g. Web Figure 1).

@@

> All analyses in the paper were conducted using a single CPU per task without multithreading. Figures 4E and 3E now exhibit the peak memory comsumption by each method along with the total computing time. We observed that in broad marks (500bp) our method was among the most efficient ones, requiring only 5GB of memory and 30 min for a genome-wide analysis of 4 samples (2 conditions, 2 replicates per condition). In narrow marks (250bp), our method required nearly 10GB of memory for a computation time of about 1.5 hour. We believe that these are reasonable quantities and quite comparable to other efficient algorithms. For example, the widely used method DiffBind required nearly as much memory even though it is a method that relies on peaks called a priori (MACS2 in our analysis) and had a much lower performance in broad histone modification marks.

@@

> As mentioned in our paper, we added a discussion in Web Appendix regarding the model selection via BIC. Web Figures 8-10 include results from our simulation study exploring the presented strategy to choose the optimal number of combinatorial patterns via BIC. Due to the lack of space, we have added the following sentence to our Discussion "When certain differential combinatorial patterns are rare, our model selection scheme via BIC provides means of pruning specific patterns of interest (Web Appendix B2).".

@@

> Input control samples were utilized by the methods that are designed and capable to do so. These methods include ChIPComp, THOR, diffReps, and DiffBind. We added the following sentence to the second paragraph of Section 5 of our paper: "Results presented here pertain to the analysis of cell lines Helas3 and Hepg2 (see Web Appendix D3 for additional results). Analyses were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR).". Similarly, input controls were also generated in our simulation study (Sequencing Read Simulation Section) and accounted for in these algorithms. Throughout our analysis, we did not observe any significant improvement in performance in epigraHMM by accounting for input control effect in differential peak calling (See Web Figure 2 for these results). This effect was also reported by others (Lun \& Smyth, 2015), in which the authors note that 'These controls are mostly irrelevant when testing for DB between ChIP samples' (see http://bioconductor.org/books/3.13/csawBook/chap-filter.html#with-negative-controls). We did not compare in our paper the performance of other methods with and without accounting for input controls. As stated above, analyses reported in our paper were adjusted by input control samples in methods that are designed to do so (ChIPComp, DiffBind, diffReps, and THOR). For all other methods (epigraHMM, RSEG, and csaw), input controls were not considered and their results are reported as such.

@@

> In Figure 2A, the observed FDR is shown for all methods and for all assessed nominal FDR levels. The observed FDR shown in Figure 2A is, on average, very close to the nominal FDR chosen (note vertical lines on the panel and corresponding nominal FDR points). Note that, overall, epigraHMM is the method that has the highest sensitivity with observed FDR closest to the nominal FDR. In Figure 3A, no FDR was or is currently reported. Instead, we show 1-specificity on the x-axis. Instead, the observed FDR is reported in Figure 3B for all methods. As we can see, the observed and nominal FDR levels reported by our method were quite close to each other (0.05, 0.07, 0.08, 0.09, 0.10). In comparison, RSEG exhibited observed FDR levels of 0.23, 0.63, 0.69, 0.74,0.80 for nominal FDR levels of 0.01, 0.05, 0.10, 0.15, and 0.20. We thank the reviewer for this comment but we do believe our results are superior than the suggested FDR control of RSEG.

@@

> We thank the reviewer for the constructive comment. We prepared a report to support our response to this comment, which can be assessed at https://github.com/plbaldoni/epigraHMMPaper/blob/main/Supp/Report_RNA_vs_ChIP/report.pdf

> As we replied to the first anonymous reviewer, we did not observe sufficient agreement between gene expression levels for the analyzed cell lines in our paper (see Figures 1 and 2 above). In Figure 2, we have an example of a H3K36me3 differentiallly enriched region and a not DE gene. In contrast, Figure 3 below presents an additional example of a DE gene and without H3K36me3 enrichment in both Helas3 and Hepg2 cell lines. Moreover, not all the analyzed marks in this paper are associated with the expression level of the nearest gene. For instance, H3K27me3 and EZH2 are two epigenomic marks known to exhibit read enrichment in intergenic regions without any nearby gene body. In several instances, enrichment of the repressive mark H3K27me3 occurs on inter-genic regions with nearby expressed genes. Figure 4 below shows an example of such a region. Due to these reasons, we believe that the proposed analysis would be less optimal for benchmarking purposes to the one presented in this paper for the analyzed cell lines. The linked report provides several other problematic cases with the peak call tracks and the entire RNA-seq data analysis.

@@

> As we discuss in our response to comment 7 of the first reviewer, Figure 5E of our paper now presents an analysis of known marker genes for cancer cell line Helas3. Specifically, we show that all known marker genes of Helas3 had associated differential combinatorial pattern associated with differential enrichment for H3K36me3 alone in our embedded mixture model. The presented differential enrichment level shown in Figure 5E is the average mixture model posterior probability associated with the mixture component that model the enrichment of H3K36me3 alone. These results highlight the benefit of the chosen modeling approach that allows not only the detection of differential regions but also the correctly classification of differential enrichment pattern, in this case the gene transcription level associated with H3K36me3. We have added the following paragraph to our paper "In fact, nearly all known Helas3 cycle-regulated
genes (therefore genes expressed at some stage of the cell life cycle; Whitfield et al. (2002)) were associated with high average mixture posterior probabilities of differential enrichment
for H3K36me3 alone (Figure 5E). The expression of such genes are known to correlate with proliferative states of tumors and their study may help characterize their role in cancer (Whitfield et al., 2002)".
